---
name: "HW 3 Assignment"
author: "Anshit Verma"
date: "12/09/2022"
output: "html_document"
---

```{r}
library(grid)
```


### Question 6.2

```{r}
data = read.table('../temps.txt', sep="\t", dec=".", header = TRUE)
data
```


```{r}
summary(data$X1996)
sd(data$X1996)
```

```{r}
mean = 83.72
st = 0
x = 8
st_df = data.frame(i=double(), st=double())
for(i in 1:nrow(data)){
  st = max(0, st + (mean - data$X1996[i] - x))
  st_df[nrow(st_df)+1,] = c(i, st)
}
plot(st_df, xlab='Day', ylab='St', type='o')
```

```{r}
yearly_mean = data.frame(DAY=data[,1], YearlyMean=rowMeans(data[,-1]))
yearly_mean
```

```{r}
summary(yearly_mean$YearlyMean)
```



```{r}
mean = mean(yearly_mean$YearlyMean)
st = 0
x = 8
st_df = data.frame(i=double(), st=double())
for(i in 1:nrow(yearly_mean)){
  st = max(0, st + (mean - yearly_mean$YearlyMean[i] - x))
  st_df[nrow(st_df)+1,] = c(i, st)
}
plot(st_df, xlab='Day', ylab='St', type='o')
```

# Inferences - 
1. Comparing the above plot with 1996, days when the temp is rising has shifted towards later part of the year, it suggests a change in season cycle.
2. Our max value of our running sum has gotten lower which also suggests warm season in later year.
3. Range of temp values has decreased (min has inc by 8 units) which has shifted temp towards mean.

# TODO - 
1. C and T hyperparameter - how to decide on values.
2. Dividing in buckets to see better changes in YoY temps.
Reference - https://rpubs.com/ssufian/658130


```{r}
### Transform to celsius
temp_cols = names(data)[!(names(data) %in% c("DAY"))]
#Transform to celsius
data_celsius = data
data_celsius[temp_cols] <- lapply(data_celsius[temp_cols], function(f) (f-32)*5/9)
```


```{r}
CUSUM_ad = function(x, year, days = data$DAY ,C = 0.30, T = 4, if_c_relative = FALSE){
  mean_x = mean(x)
  sd_x = sd(x)
  if (if_c_relative) {
    C = C * sd_x
  }
  #apply cusum
  x1 = lapply(x, function(xi)  (mean_x - xi - C))
  st_df = data.frame(st=double())
  st = 0
  for(i in 1:length(x)){
    #print (paste(i,class(st),class(x1[[i]])))
    st = max(0, st + x1[[i]])
    st_df[i,] = c(st)
  }
  #Find index where St > T
  st_df$Day = days
  st_df["anamoly"] = 0
  st_df$anamoly[st_df$st > T] = 1
  st_df$xt = x
  #plot x with T and if decrease see
  par(mfrow=c(2,1))
  plot(st_df$x)
  title(paste("Temperature (in Celsius) and S_t (bottom) for year",year), line = -1, outer = TRUE)
  plot(st_df$st)
  pushViewport(viewport())
  grid.lines(x = c(0,1), y = grconvertY(T, "user", "ndc"), gp = gpar(col = "red"))
  change_detected = which.max(st_df$st>T)
  grid.lines(x = grconvertX(change_detected, "user", "ndc"), y = c(0,1), gp = gpar(col = "red"))
  popViewport()
  #return first point where St touches T
  change_length = length(days)-change_detected
  return (c(as.numeric(year),as.numeric(change_detected),as.numeric(change_length)))
}
```



```{r}
### Run Cusum over data_celsius for all the years
winter_start_days = lapply(temp_cols, function(col) 
  CUSUM_ad(x=data_celsius[,col], year = substr(col,2,5), C = 4, T = 25)
  )
```

```{r}
change_df <- as.data.frame(do.call(rbind, winter_start_days))
colnames(change_df) = c("year","winter_start","winter_length")
change_df["year_bucket"] = "a.1996_2000"
change_df[change_df$year>=2000,"year_bucket"] = "b.2000_2004"
change_df[change_df$year>=2004,"year_bucket"] = "c.2004_2008"
change_df[change_df$year>=2008,"year_bucket"] = "d.2008_2012"
change_df[change_df$year>=2012,"year_bucket"] = "e.2012_2016"
change_df
```

```{r}
aggregate(cbind(winter_start,winter_length) ~ year_bucket, change_df, mean)
```


```{r}
month_df <- data.frame(data)
month_df["month"] = "month"
month_df[grepl( "Jul", month_df$DAY), "month"] = "Jul"
month_df[grepl( "Aug", month_df$DAY), "month"] = "Aug"
month_df[grepl( "Sep", month_df$DAY), "month"] = "Sep"
month_df[grepl( "Oct", month_df$DAY), "month"] = "Oct"
month_df = month_df[,-1]
for ( col in 1:ncol(month_df)){
    colnames(month_df)[col] <-  sub("X", "", colnames(month_df)[col])
}
month_df
```

```{r}
year_bucketing_function = function(df){
  df["year_bucket"] = "a.1996_2000"
  df[df$Year>=2000,"year_bucket"] = "b.2000_2004"
  df[df$Year>=2004,"year_bucket"] = "c.2004_2008"
  df[df$Year>=2008,"year_bucket"] = "d.2008_2012"
  df[df$Year>=2012,"year_bucket"] = "e.2012_2016"
  df
}
```


```{r}
aggregate_on_month_and_transpose = function(df, func){
  return_df = aggregate(. ~ month, df, func)
  n = return_df$month
  # transpose all but the first column (month)
  return_df <- as.data.frame(t(return_df[,-1]))
  colnames(return_df) <- n
  return_df = cbind(Year = rownames(return_df), return_df)
  rownames(return_df) = 1:nrow(return_df)
  return_df = return_df[, c(1, 3, 2, 5, 4)]
  return_df
}
```



```{r}
mean_df = aggregate_on_month_and_transpose(month_df, mean)
mean_df = year_bucketing_function(mean_df)
mean_df = mean_df[,-1]
mean_df = aggregate(.~year_bucket, mean_df, mean)
mean_df
```

```{r}
min_df = aggregate_on_month_and_transpose(month_df, min)
min_df = year_bucketing_function(min_df)
min_df = min_df[,-1]
min_df = aggregate(.~year_bucket, min_df, mean)
min_df
```


```{r}
max_df = aggregate_on_month_and_transpose(month_df, max)
max_df = year_bucketing_function(max_df)
max_df = max_df[,-1]
max_df = aggregate(.~year_bucket, max_df, mean)
max_df
```


