---
title: "ad_ques12.2"
author: "Ashish Dhiman"
date: "2022-10-19"
output: html_document
---

```{r}
library(FrF2)
```

## Question 12.2

We want to design a experiment such that we can find the effect of 10 Binary features on the price of a house. Some of the sample features could be:

1.  House has more than 4 bedrooms
2.  Park nearby
3.  School nearby
4.  If Pets are allowed in Locality, etc

With the full factorial design, we would land with $$2^{10} = 1024$$ combinations, which is a very large number.

Hence we want to design a fractional factorial design experiment here, with 16 combinations.

```{r}
n_combinations = 16
n_variables = 10
```

```{r}
factorial_doe = FrF2(nruns = n_combinations,nfactors =  n_variables,
                     default.levels = c(0,1))
factorial_doe
```

Note here 1 means binary variable present, and 0 means otherwise

```{r}
summary(factorial_doe)
```

The #generators =10-4 = 6. B'cos 16 = 2\^4.

### Properties of Factorial Design

#### Each choice should be checked same number of times

```{r}
table(factorial_doe$A)
```

```{r}
print (paste("Shape of original df: ",dim(factorial_doe)[1]))
print (paste("Shape of de-deuplicated df: ",dim(unique(as.data.frame(factorial_doe)))[1]))
```

The above test shows that the table is unique without any duplicates. And, since A is tested only 8 times, the information can be combined to assert the above test.

#### All pair of choices should be chosen the same number of times

```{r}
table(factorial_doe$A,factorial_doe$B)
```

```{r}
table(factorial_doe$A,factorial_doe$D)
```

Above Behavior can again be generalized to prove the second assumption.

### Feature distribution

```{r}
factorial_doe1 = as.data.frame(factorial_doe)
```

```{r}
factorial_doe1$sum = as.numeric(factorial_doe1$A)+as.numeric(factorial_doe1$B)+as.numeric(factorial_doe1$C)+as.numeric(factorial_doe1$D)+as.numeric(factorial_doe1$E)+as.numeric(factorial_doe1$F)+as.numeric(factorial_doe1$G)+as.numeric(factorial_doe1$H)+as.numeric(factorial_doe1$J)+as.numeric(factorial_doe1$K)-10
table(factorial_doe1$sum)
```
