---
title: "ISYE 6501"
output:
  html_document:
    df_print: paged
  pdf_document: default
date: "2022-09-12"
---

# Submission HW8 \| Fall 22

-   Ashish Dhiman \| [ashish.dhiman\@gatech.edu](mailto:ashish.dhiman@gatech.edu){.email}

-   Abhinav Arun \| [aarun60\@gatech.edu](mailto:aarun60gatech.edu){.email}

-   Anshit Verma \| [averma373\@gatech.edu](mailto:averma373@gatech.edu){.email}

## Question 12.1

Design of Experiments finds its use cases in various domains :

One particular area where it is widely used is in : 

***Website design prototyping and Call to Action (CTA) Optimization***


We can use it to evaluate the impact of adding a new functionality on a website to evaluate user interaction/user traffic on a website : 

There are various aspects which we could experiment on : 

* Location of the new feature on the website : 4 possible categories (Top Left , Top Right , Bottom Left , Bottom Right) to evaluate whether the new feature location would increase user engagement.

* Color of the feature added : 3 possible categories (red,blue,green) - to see if a the color of the feature affects user interaction

* Feature mobility: 2 categories (dynamic, static) - whether the feature should be static or dynamic (can help us evaluate if people are liking simplistic feature or they like more interactive/dynamic features).

* Having couple of icons on the new feature : One good use case of this is when creating a "Pay Now" icon on the payment gateway website . Should we add an additional icon of "Return Back" adjacent to Pay now button because it may so happen that many people might not go through payment transactions and as a result of it the company's profitability might get impacted.So it's better to test out it's impact with some real data (Yes/No option , so 2 possible categories).

* Size of the feature added : 3 possible categories (Small, Medium, Large) and it's impact on user liking/interaction.

Thus , we will have a total of $4\cdot 3\cdot 2\cdot 2 \cdot 3$ = 144 combinations to test out . In order to get a better sampling distribution across each combination , we can think on reducing the total number of combinations using various factorial design techniques so that we do not have very less data corresponding to a possible combination and so that we could make some statistically significant inference about the user behavior.


## Question 12.2

```{r, include=FALSE}
library(FrF2)
```

We want to design a experiment such that we can find the effect of 10 Binary features on the price of a house. Some of the sample features could be:

1.  House has more than 4 bedrooms
2.  Park nearby
3.  School nearby
4.  If Pets are allowed in Locality, etc

With the full factorial design, we would land with $$2^{10} = 1024$$ combinations, which is a very large number.

Hence we want to design a fractional factorial design experiment here, with 16 combinations.

```{r}
n_combinations = 16
n_variables = 10
```

```{r}
factorial_doe = FrF2(nruns = n_combinations,nfactors =  n_variables,
                     default.levels = c(0,1))
factorial_doe
```

Note here 1 means binary variable present, and 0 means otherwise

```{r}
summary(factorial_doe)
```

The #generators =10-4 = 6. B'cos 16 = 2\^4.

### Properties of Factorial Design

#### Each choice should be checked same number of times

```{r}
table(factorial_doe$A)
```

```{r}
print (paste("Shape of original df: ",dim(factorial_doe)[1]))
print (paste("Shape of de-deuplicated df: ",dim(unique(as.data.frame(factorial_doe)))[1]))
```

The above test shows that the table is unique without any duplicates. And, since A is tested only 8 times, the information can be combined to assert the above test.

#### All pair of choices should be chosen the same number of times

```{r}
table(factorial_doe$A,factorial_doe$B)
```

```{r}
table(factorial_doe$A,factorial_doe$D)
```

Above Behavior can again be generalized to prove the second assumption.

### Feature distribution

```{r}
factorial_doe1 = as.data.frame(factorial_doe)
```

```{r}
factorial_doe1$sum = as.numeric(factorial_doe1$A)+as.numeric(factorial_doe1$B)+as.numeric(factorial_doe1$C)+as.numeric(factorial_doe1$D)+as.numeric(factorial_doe1$E)+as.numeric(factorial_doe1$F)+as.numeric(factorial_doe1$G)+as.numeric(factorial_doe1$H)+as.numeric(factorial_doe1$J)+as.numeric(factorial_doe1$K)-10
table(factorial_doe1$sum)
```


## Question 13.1

Examples of data for the following distributions:

a. **Binomial** - In finding defective items, the outcome could be either success (item is defective) or failure (item is non-defective). If you are doing this for multiple items and each item being selected is at random i.e. iid, then this example follows a Binomial distribution.
b. **Geometric** - For example, you ask people outside a polling station who they voted for until you find someone that voted for the independent candidate in a local election. The geometric distribution would represent the number of people who you had to poll before you found someone who voted independent. You would need to get a certain number of failures before you got your first success. This example follows a Geometric distribution.
c. **Poisson** - It gives the probability of an event happening a certain number of times (k) within a given interval of time or space. One example is no. of website visitors per month. We are using the arrivals of vistors to a website on a per month basis. This follows a Poisson distribution.
d. **Exponential** - One example is the amount of time until an earthquake occurs. The time interval between earthquake follows an exponential distribution.
e. **Weibull** - One example is lifetime of dental implants. We are using data to model time-to-failure. If you get new dental implants, after how long will the implants breaks. This follows a Weibull distribution.
